---
title: "How to visualise data in R"
date: last-modified
date-format: "[Last updated:] DD/MM/YYYY HH:mm"
format:
  html:
    standalone: true
    embed-resources: true
    toc: true
    toc-depth: 3
    toc-expand: 1
    number-sections: true
    toc-location: left
    html-table-processing: none
    df-print: kable
    code-fold: false
    theme: default
    fontsize: 1em
    link-external-newwindow: true
execute:
  warning: false
  echo: true
  eval: true
---

A compendium of code for visualising data in R (primarily using `ggplot2`).

```{r}
#| label: packages
#| echo: false
#| warning: false
#| message: false
library(pacman)
p_load(
  tidyverse,
  janitor,
  gcookbook,
  palmerpenguins,
  scales,
  RColorBrewer,
  networkD3,
  purrr,
  ggExtra,
  grid,
  gridExtra,
  GGally,
  ggpubr,
  directlabels,
  ggrepel,
  gghighlight,
  afcharts,
  gapminder,
  devtools)
```

Note: this guide uses built-in datasets from the following, amongst others:

- [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/)
- [`gcookbook`](https://cran.r-project.org/web/packages/gcookbook/gcookbook.pdf)
- [`ggplot2`](https://ggplot2.tidyverse.org/reference/#data)
- [`gapminder`](https://cran.r-project.org/web/packages/gapminder/readme/README.html)

# Structure of a `ggplot`

* **Aesthetics** describe mapping between visual elements and variables in the data, e.g. x-axis may be mapped to "time_point", while colour may be mapped to "gender". 
* **Geoms** are the type of visual 'marks' on a plot such as lines, points, or bars (covered under Plot types below): they are geometrical objects used to represent data.

```{r}
#| eval: false
data %>% 
  ggplot(aes(x = ind_var, 
             y = dep_var)) +
  geom_point(aes(colour = factor(grouping_var)),
             size = 1.5)
```

If you want to add a global attribute (e.g. to apply to all points, lines, or whatever), specify this **outside of `aes()`** because it is not a mapping (it doesn't relate to something in the dataframe itself). 

# Plot types
## Points
```{r}
#| fig-height: 3
#| fig-width: 5
mpg %>%
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(size = 1.5,
             position = "identity")
```

This masks the picture because there are overlapping points. Changing `position` to `"jitter"`:
```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 5
mpg %>%
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(size = 1.5,
             position = "jitter") 
```

Adding mappings for colour and shape:
```{r}
#| fig-height: 3
#| fig-width: 7
mpg %>%
  ggplot(aes(x = displ, y = hwy)) +
  geom_point(aes(colour = class,
                 shape = class),
             size = 1.5,
             position = "jitter") 
```

Note that this can also be written as:
```{r}
#| eval: false
mpg %>%
  ggplot() +
  geom_point(aes(x = displ, y = hwy, 
                 colour = class,
                 shape = class), 
             size = 1.5)
```

### Point shapes
Note: the `fill` attribute can be changed to any colour for shapes 21 to 24. 
```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 3

shapes <- data.frame(
  shape = c(0:19, 22, 21, 24, 23, 20),
  x = 0:24 %/% 5,
  y = -(0:24 %% 5))

ggplot(shapes, aes(x, y)) + 
  geom_point(aes(shape = shape), size = 5, fill = "red") +
  geom_text(aes(label = shape), hjust = 0, nudge_x = 0.15,
            show.legend = T) +
  scale_shape_identity() +
  expand_limits(x = 4.1) +
  theme_void()
```

## Lines
```{r}
#| fig-height: 3
#| fig-width: 6
gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "Ireland") & Year > 1980) %>%
  ggplot(aes(x = Year, y = GDP)) +
  geom_line(aes(colour = Name, 
                linetype = Name), 
            linewidth = 0.8)

# to add arrows to the ends of lines, add this to geom_line:
# arrow = arrow(length = unit(0.25, "cm"), ends = "last", type = "closed")
```

Adding a line of best fit
```{r}
#| fig-height: 3
#| fig-width: 5
gcookbook::heightweight %>%
  ggplot(aes(x = heightIn, y = weightLb)) +
  geom_point(aes(colour = sex)) +
  geom_smooth(method = "lm", # or "loess"
              fullrange = T,
              aes(colour=sex))
```


Use `geom_segment` to add lines from points to fitted regression slope using some dummy data:
```{r}
#| echo: false
# prepare data
x = c(12,	3,	5,	3,	1,	15,	18,	10,	8,	3)
y = c(10,	2,	4,	9,	5,	13,	14,	11,	12,	4)
df = data.frame(x, y)

mod <- lm(y ~ x, data = df)
df <- transform(df, Fitted = fitted(mod))
head(df)
```


```{r}
#| fig-height: 3
#| fig-width: 5

ggplot(df, aes(x = x, y = y)) + 
  geom_point() + 
  theme_classic() + 
  geom_smooth(method = "lm", se = F) + 
  geom_segment(aes(x = x, y = y,
                   xend = x, yend = Fitted), 
               linetype = "dashed")
```


### Linetypes
Example usage: `linetype = "dashed"`
```{r}
#| echo: false
#| fig-height: 3
#| fig-width: 5
lty <- c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash")
linetypes <- data.frame(
  y = seq_along(lty),
  lty = lty) 

ggplot(linetypes, aes(0, y)) + 
  geom_segment(aes(xend = 5, yend = y, linetype = lty)) + 
  scale_linetype_identity() + 
  geom_text(aes(label = lty), hjust = 0, nudge_y = 0.2) +
  scale_x_continuous(NULL, breaks = NULL) + 
  scale_y_reverse(NULL, breaks = NULL)
```


## Bars
### Simple bar chart

* `geom_col` leaves the data as it is and merely represents values already in the dataframe.
* `geom_bar` uses `stat_count` by default to derive new values from the data. As a result, `geom_bar` doesn't expect a y-value, but if you provide one then you are telling it to forgo the aggregation it would have done anyway with `stat_count`.

Using `stat = "count"`
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = species, fill = species)) +
  geom_bar(stat = "count", # default - can be omitted
           width = 0.8,
           show.legend = F)
```

Using `stat = "identity"` to represent pre-calculated summary stats.
```{r}
#| fig-height: 3
#| fig-width: 4
gcookbook::drunk %>%
  pivot_longer(c(2:6)) %>%
  group_by(sex) %>%
  summarise(felonies = sum(value)) %>%
  ggplot(aes(x = sex, y = felonies)) +
  geom_bar(stat = "identity", width = 0.8)
```

You should also use `stat_identity` if you want to **reorder** bars in descending order. The syntax for `reorder` is `reorder(what, by_what)`. Here, `-n` means descending order. 
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  group_by(species) %>%
  summarise(n = n()) %>%
  ggplot(aes(reorder(species, -n), n)) +
  geom_bar(stat = "identity", width = 0.8, 
           aes(fill = species),
           show.legend = F)
```


### Dodged bar chart
Dodged bars represent geoms for each level of some factor (here, sex), arranged side by side for ease of comparison. Note that this dataset has NAs. To remove these before graphing, use `drop_na`.
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  # drop_na(sex) %>%
  ggplot(aes(x = island, fill = sex)) +
  geom_bar(position = position_dodge(),
           width = 0.8)
```

### Stacked bar chart
Stacked bars are placed on top of one another. Here is a recipe for when you already have a y-axis (e.g. counts or figures for each category of something).
```{r}
#| fig-height: 3
#| fig-width: 5

palmerpenguins::penguins %>%
  ggplot(aes(x = island, fill = sex)) +
  geom_bar(position = position_stack(), 
           width = 0.8)
```


## Proportional stacked bar
### When there is an explicit y-value
Use this recipe if your dataframe already has a variable representing the y-value (e.g. `count`). Notice use of `stat_identity`.
```{r}
#| echo: false

prop_df = data.frame(
  group = c("a", "a", "b", "b"),
  type = c("x", "y", "x", "y"),
  count = c(11091, 4583, 3974, 10984)
)
```

```{r}
#| fig-height: 3
#| fig-width: 5

prop_df

prop_df %>%
  ggplot(aes(x = group, y = count, fill = type)) +
  geom_bar(position = position_fill(), stat = "identity", width = 0.8) +
  ylab("Proportion")
```

### When there is no explicit y-value (i.e. counts of a factor)
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = island, fill = sex)) +
  geom_bar(position = position_fill(),
           width = 0.8) +
  ylab("Proportion")
```

Text is dealt with later on, but to add figures to the above figure you need to manually calculate proportions for each cell first and then pipe this into ggplot. The recipe below also has some nifty code for converting decimals into nicely formatted percentages.
```{r}
#| fig-height: 3.5
#| fig-width: 5
palmerpenguins::penguins %>%
  group_by(island, sex) %>%
  summarise(n = n()) %>% # you may need to use sum() here instead of n()
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = island, y = prop, fill = sex)) +
  geom_bar(stat = "identity", width = 0.8) +
  geom_text(aes(label = paste0(sprintf("%.1f", prop * 100), "%")), 
            position = position_stack(vjust = 0.5),
            size = 3,
            show.legend = F) +
  ylab("Proportion")
  
# note: to change decimal places, "%.0f", "%.2f" etc
```

## Plotting summary data
Use `stat_summary` to plot statistical summaries like sums or means. This requires raw-format data (i.e. which hasn't already been summarised) as `ggplot` does the calculations internally. See @sec-summarydata for more information. 

```{r}
#| echo: false
df = data.frame(
  year = c(2011, 2011, 2011, 2012, 2012, 2012, 2013, 2013, 2013, 2014, 2014, 2014),
  value = c(10, 15, 10, 20, 10, 10, 15, 30, 20, 10, 20, 10)
)
```

```{r}
#| fig-height: 3
#| fig-width: 5
df %>%
  ggplot(aes(x=year, y=value)) +
  stat_summary(fun = "sum", geom = "line")  +
  ggtitle("fun = 'sum', geom = 'line'")

df %>%
  ggplot(aes(x=year, y=value)) +
  stat_summary(fun = "mean", geom = "bar") +
  ggtitle("fun = 'mean', geom = 'bar'")
```

## Boxplot
```{r}
#| fig-height: 3
#| fig-width: 5

palmerpenguins::penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  # this part produces just the I-shaped lines
  stat_boxplot(
    geom ='errorbar', 
    width = 0.5) +
  # this part produces the boxes
  geom_boxplot(
    notch = F,
    outlier.color = "red",
    outlier.size = 2.5
  )
```

## Violin plot
A more sophisticated version of a boxplot using density curves to show the distribution for the entire range of values. 
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = species, y = bill_length_mm)) +
  geom_violin(aes(fill = species),
              draw_quantiles = c(0.5)) # specify 0.5 for median
```

## Dotplot
For when there isn't a huge amount of data. Each dot represents a single observation.
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  sample_n(125, replace=F) %>%
  ggplot(aes(x=body_mass_g)) +
  geom_dotplot(dotsize = 1, width = 1)
```


## Histogram
For when there's more data, or binning is better. This only requires an x-axis value. `geom_histogram` usually does a good job of setting the `bindwidth` automatically but this can be manually controlled. 
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = body_mass_g)) +
  geom_histogram(binwidth = 100, 
                 fill = "grey", 
                 colour = "black")
```

Recipe: overlay separate histograms for levels of a factor (here, `cut`). By default, `geom_histogram` will stack bars if there are multiple groups but you can change this by changing `position` to `identity`:
```{r}
#| fig-height: 3
#| fig-width: 6
ggplot2::diamonds %>%
  ggplot(aes(x = carat, fill = cut)) +
  geom_histogram(binwidth = 0.1, position = "identity", alpha = 0.5)
```

Sometimes you may want to produce a range of histograms for each level of a factor. Here is a recipe for a function to iterate through multiple histograms, by group, using `purrr`:
```{r}
#| fig-height: 3
#| fig-width: 4
# firstly make a vector of all the variable names you want to plot
vars = c("bill_length_mm", "flipper_length_mm", "body_mass_g")

# then create a custom graphing function
hist_fun = function(data, x, y) {
  ggplot(data, aes(x = .data[[x]], fill = .data[[y]]) ) +
    geom_histogram(alpha = 0.5, position = "identity") +
    theme_bw() +
    ggtitle(x)
}

# use purrr::map to cycle through vars to produce plots, with a constant grouping factor
purrr::map(vars, ~ hist_fun(data = palmerpenguins::penguins, .x, "sex") )
```

## Density
For large amounts of data. Use `bw` argument to change bandwidth (e.g. if you want more smoothing).
```{r}
#| fig-height: 3
#| fig-width: 5
ggplot2::diamonds %>%
  ggplot(aes(x = carat)) +
  geom_density(bw = 0.05, fill = "lightblue", 
               alpha = 0.8)
```

As with the multiple-group histogram example above, you may want to overlay density distributions for levels of a factor.
```{r}
#| fig-height: 3
#| fig-width: 5
ggplot2::diamonds %>%
  ggplot(aes(x = carat, fill = cut)) +
  geom_density(bw = 0.05, alpha = 0.4)
```



## Pie
Everyone knows that pie charts are illegal in the world of data visualisation, but here's a recipe just in case...
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  count(species) %>%
  ggplot(aes(x = "", y = n, fill = species)) +
  geom_col(color = "black") +
  coord_polar(theta = "y") +
  geom_text(aes(label = n),
            position = position_stack(vjust = 0.5),
            show.legend = F) +
  theme_void()
```

## Sankey
This recipe produces an interactive html Sankey chart using the `networkD3` package. It may take a little set up beforehand according to the data you have. 
```{r}
#| fig-height: 4
#| fig-width: 5

links = data.frame(
  Var1 = c("Failed_Phonics", "Passed_Phonics", "Failed_Phonics", "Passed_Phonics"),
  Var2 = c("Failed_KS2", "Failed_KS2", "Passed_KS2", "Passed_KS2"),
  Freq = c(25114, 120356, 9019, 585893)
)

colnames(links) <- c("source", "target", "value")

nodes <- data.frame(name = as.factor(c("Failed_Phonics","Passed_Phonics","Failed_KS2","Passed_KS2")))

# Convert the source and target from characters to indices
links$source <- match(links$source, nodes$name) - 1
links$target <- match(links$target, nodes$name) - 1

# create Sankey diagram
networkD3::sankeyNetwork(Links = links, 
              Nodes = nodes, 
              Source = "source",
              Target = "target", 
              Value = "value", 
              NodeID = "name", 
              fontSize = 12, 
              fontFamily = "sans-serif")
```

This can be made into a function to be applied iteratively. 
```{r}
#| eval: false
# function
sankey_function = function(var1, var2, lab1, lab2, lab3, lab4){
  
  tab = table(var1, var2)
  
  links = data.frame(
    Var1 = c(lab1, lab2, lab1, lab2),
    Var2 = c(lab3, lab3, lab4, lab4),
    Freq = c(tab[1,1], tab[2,1], tab[1,2], tab[2,2])
  )
  
  colnames(links) <- c("source", "target", "value")
  
  nodes <- data.frame(name = as.factor(c(lab1, lab2, lab3, lab4)))
  
  links$source <- match(links$source, nodes$name) - 1
  links$target <- match(links$target, nodes$name) - 1
  
  # create Sankey diagram
  networkD3::sankeyNetwork(Links = links, 
                           Nodes = nodes, 
                           Source = "source",
                           Target = "target", 
                           Value = "value", 
                           NodeID = "name", 
                           fontSize = 12, 
                           fontFamily = "sans-serif")
}


sankey_function(var1 = ela_final$ks4_mfl, var2 = ela_final$ks5_mfl_any,
                lab1 = "ks4_mfl_no", lab2 = "ks4_mfl_yes",
                lab3 = "ks5_mfl_no", lab4 = "ks5_mfl_yes")
```


# Some more advanced recipes
Add a histogram to a scatterplot using `ggExtra::ggMarginal`. This requires you to set up the basic plot first and save as an object. 
```{r}
#| fig-height: 3
#| fig-width: 4

# create data
set.seed(30)
df1 = data.frame(x = rnorm(500, 50, 10), y = runif(500, 0, 50))
# set up base plot
p1 = ggplot(df1, aes(x, y)) + geom_point() + theme_bw()

ggMarginal(p1, type = "histogram",
           margins = "both", # note: "x", "y", or "both" 
           fill = "orange",
           xparams = list(binwidth = 2))
```


Create a scatterplot matrix using `GGally::ggpairs`. This is quite customisable depending on what you need. Check `help(ggpairs)`
```{r}
#| message: false
#| fig-height: 7
# note: you can also add colouring by factor if you add: ggplot2::aes(colour = Species)
GGally::ggpairs(iris)
```


# Plotting statistical summary data {#sec-summarydata}
Often we want to plot statistical summaries rather than raw data. `ggplot` has functions to do this "in-plot", i.e. instead of you having to compute summaries manually beforehand. One option is to use arguments within `geom_bar`.
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = species, y = flipper_length_mm)) +
  geom_bar(fun = "median", 
           stat = "summary")
```

However, a more powerful method is to use `stat_summary`. You can plot several summaries together, e.g. if you want points and lines on the same plot. 
```{r}
#| fig-height: 3
#| fig-width: 5
palmerpenguins::penguins %>%
  ggplot(aes(x = species, y = flipper_length_mm)) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line", aes(group = 1)) 
```

Plot the mean and standard deviation using `mean_sdl` (use the `mult` argument to change how many standard deviations are shown around the mean):
```{r}
#| fig-height: 3
#| fig-width: 5
mpg %>%
  ggplot(aes(x = reorder(class, hwy), y = hwy)) +
  stat_summary(fun = mean, geom = "point") +
  stat_summary(fun.data = mean_sdl,
               fun.args = list(mult = 1),
               geom = "errorbar",
               width = .4)
```

Plot the interquartile range (1st to 4th quartiles with median):
```{r}
# custom function
median_IQR <- function(x) {
  data.frame(y = median(x), # Median
             ymin = quantile(x)[2], # 1st quartile
             ymax = quantile(x)[4])  # 3rd quartile
}

mpg %>%
  ggplot(aes(x = class, y = hwy)) +
  stat_summary(geom = "linerange",
               fun.data = median_IQR, 
               linewidth = 1.5) +
  stat_summary(fun = median, geom = "point", colour = "red", size = 2.5)
```
```{r}
# Another option if you want to plot manually created min and max values, and add
# a separate variable as well as the linerange:
# NOTE: you must wrap quantile() in as.numeric() otherwise this will produce a
# NAMED vector instead of scalar values which can cause issues when rending Quarto.
data = airquality %>%
  group_by(Month) %>%
  summarise(q1 = as.numeric(quantile(Temp, na.rm=T, probs=0.25)),
            median = median(Temp, na.rm=T),
            q3 = as.numeric(quantile(Temp, na.rm=T, probs=0.75)),
            ozone = median(Ozone, na.rm=T))

data %>%
  ggplot(aes(x = reorder(Month, median), y = median)) +
  geom_linerange(aes(ymin = q1, ymax = q3)) +
  geom_line(aes(y = ozone, colour = "ozone", group = 1)) +
  theme_classic()
```

You could also use `geom_ribbon` instead (this time there's a grouping factor and the median is superimposed in the ribbons too):
```{r}
data2 = airquality %>%
  mutate(group = rep(c("a", "b"), length.out = n())) %>%
  group_by(Month, group) %>%
  summarise(q1 = as.numeric(quantile(Temp, na.rm=T, probs=0.25)),
            median = median(Temp, na.rm=T),
            q3 = as.numeric(quantile(Temp, na.rm=T, probs=0.75)),
            ozone = median(Ozone, na.rm=T)) %>%
  mutate(q1 = if_else(group == "a", q1+5, q1),
         q3 = if_else(group == "a", q3+5, q3),
         median = if_else(group == "a", median+5, median))

data2 %>%
  ggplot(aes(x = Month)) +
  geom_ribbon(data = data2, aes(ymin = q1, ymax = q3, group = group, fill = group),
                 position = position_dodge(width = 0.1), alpha = 0.5) +
  geom_line(aes(y = median, colour = group, group = group),
            position = position_dodge(width = 0.1), linewidth = 1) +
  theme_classic() +
  scale_fill_manual(values = c("#9FB9C8", "#E87D1E"),
                        breaks = c("a", "b"),
                        labels = c("a", "b")) +
  scale_colour_manual(values = c("#9FB9C8", "#E87D1E"),
                        breaks = c("a", "b"),
                        labels = c("a", "b"))
```




Add confidence intervals (requires the `Hmisc` package):
```{r}
#| fig-height: 3
#| fig-width: 5
mpg %>%
  ggplot(aes(x = reorder(class, hwy), y = hwy)) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_cl_normal",
               fun.args = list(conf.int = .95),
               geom = "errorbar",
               width = .4) 
```



## Pointrange
This shows the mean and SE. 
```{r}
#| fig-height: 4
#| fig-width: 4

gapminder::gapminder %>%
  filter(year %in% c(1950:1990)) %>%
  ggplot() +
  geom_pointrange(mapping = aes(x = year, y = lifeExp),
                  stat = "summary")
```

# Working with text
This diagram shows the main text elements in a ggplot2 graph, all of which are controllable (see below).
```{r}
#| fig-height: 3
#| fig-width: 5
#| echo: false
ggplot(data.frame(x = c("axis.text.x"), y= c("axis.text.y")), aes(x=x, y=y)) +
  xlab("xlab/ axis.title.x") +
  ylab("ylab / \n axis.title.y") +
  ggtitle("ggtitle", subtitle = "subtitle") +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
```


## Plot title, subtitle, and caption
```{r}
#| eval: false
p + ggtitle("Main plot title", subtitle = "Plot subtitle")

# use labs to add a caption/footnote at the bottom right of the plot
p + labs(caption = "my caption")

# caption is right-justified by default. To change:
p + theme(plot.caption = element_text(hjust=0))

# centre align plot title
p + theme(plot.title = element_text(hjust = 0.5))

# change plot title size
p + theme(plot.title = element_text(size = 18))
```


## Adding text to charts
If you just want to add static text somewhere on a chart, use this:
```{r}
# for plot with numeric axes:
plot + geom_text(x = 100, y = 100, label = "my label", hjust = 1.2)

# for plot with date on x-axis
plot + geom_text(x = as.Date("2025-09-01"), y = 100, label = "my label", hjust = 1.2)
```


Note that you should use `geom_label` if you want white boxes surrounding the text. The example below calculates percentage changes for each month for 3 variables and adds text to show these percentages. Pay attention to the legend - by default it includes little 'a' characters because it's linked to text. To turn this off, specify `show_legend = F` in `geom_label` (then the legend will revert to showing dots and lines). 
```{r}
df = airquality %>%
  group_by(Month) %>%
  summarise(ozone_m = mean(Ozone, na.rm=T),
            solar_m = mean(Solar.R, na.rm=T),
            wind_m = mean(Wind, na.rm=T)) %>%
  pivot_longer(-c(Month)) %>%
  arrange(name, Month) %>%
  group_by(name) %>%
  mutate(lag = dplyr::lag(value, n = 1, default = NA)) %>%
  mutate(diff = value - lag) %>%
  mutate(change_pc = diff / lag * 100)
df

df %>%
  ggplot(aes(x = Month, y = value, group = name, colour = name)) +
  geom_line() +
  geom_point() +
  geom_label(aes(label = round(change_pc,1)), 
             size = 3.5, 
             alpha = 0.7, 
             vjust = -0.5,
             colour = "black", # if you don't want text to be coloured
             border.colour = "grey70",
             show.legend = T) +
  theme_classic() +
  coord_cartesian(ylim = c(0,250))
```



Recipe to add the stat to the top of a bar.
```{r}
#| fig-height: 4
#| fig-width: 4
palmerpenguins::penguins %>%
  group_by(species) %>%
  summarise(n = n()) %>%
  ggplot(aes(x = species, y = n)) +
  geom_bar(stat = "identity", width = 0.8) +
  geom_text(
    aes(label = n, vjust = -0.5),
    size = 3.5)
```

It's also possible to display text for certain conditions.
```{r}
#| eval: false
# p + geom_text(aes(label = ifelse(condition == "con", paste0(round(diff_pc,1), "%"), ""), vjust = -0.5),size = 3.5)

# a more common strategy is to calculate bar counts beforehand and then plug these in later.
bar_data = ela_final %>%
  filter(!is.na(status)) %>%
  # reorder the factor so that the order of filled bars is descending (first 
  # on top, last at bottom)
  mutate(status = factor(status, levels = c("Deceased", "Cancelled", "Rejected",
                                            "Waiting list", "Matching", 
                                            "Applicant withdrawn", "Allocated"))) %>%
  count(AY, status)

ggplot(bar_data, aes(x = AY, y = n, fill = status)) +
  geom_bar(stat = "identity") +
  geom_text(
    # just show counts from "Allocated" status
    data = bar_data %>% filter(status == "Allocated"),
    aes(x = AY, y = n, label = n),
    position = position_stack(vjust = 0.5),
    color = "black",
    size = 3.5
  )
```


# Axes
Note that the same functions are used for x or y axes - adjust accordingly.
```{r}
#| eval: false
# custom axis titles
p + xlab("...")
p + ylab("...")

# remove axis title
p + theme(axis.title.x = element_blank())

# change axis title text size
p + theme(axis.title.x = element_text(size = 14))

# rotate x- axis title text 45 degrees
p + theme(axis.title.x = element_text(angle = 45, vjust = 0.5))

# rotate y-axis to 0 degrees (to be read horizontally not vertically)
p + theme(axis.title.y = element_text(angle = 0, vjust = 0.5))

# remove tick marks
p + theme(axis.ticks.x = element_blank())
```


## Axis formatting options
```{r}
#| eval: false
# percent labels with breaks of 10-100%
# note: change scale to 0 if you are working with integers already, otherwise leave at default of 100
p + scale_y_continuous(labels = scales::label_percent(accuracy = 1, scale = 100), breaks = (0:10)/10)

# currency suffix
p + scale_y_continuous(labels = scales::label_dollar(prefix = "£"),
                       breaks = seq(0, 100, by = 25))
# can also specify suffix, big.mark = ",", and decimal.mark = "."

# if you just want to display all breaks without listing them:
p + scale_x_continuous(n.breaks = 17)

# thousands separator
p + scale_y_continuous(labels = scales::label_comma(big.mark = ","))

# dates and times
p + scale_y_continuous(labels = label_date(format = "%Y-%m-%d"))
p + scale_y_continuous(labels = label_time(format = "%H:%M:%S"))
```


Using `scale_x_date`: this requires the x-axis to be formatted as a date object.
```{r}
#| eval: false
# firstly, a nifty function if your date value is in the format "201516"
convert_year = function(year){
  yr = substr(year, 5, 6)
  full_yr = paste0("20", yr, "-01-01")
  full_yr_date = as.Date(full_yr, format = "%Y-%m-%d")
}

# Note that you can either supply date_labels OR a character vector of custom labels, not both. 
p + scale_x_date(date_labels = "%Y", date_breaks = "1 year")

# if your x-axis is in date format, you can try the following:
yrs = c(as.Date("2016-06-01"), as.Date("2017-06-01"), as.Date("2018-06-01"))
yrs_labs = c("15/16", "16/17", "17/18")
p + scale_x_date(breaks = yrs,
                 labels = yrs_labs)

# I've found that I get an error saying `breaks` and `labels` have different lengths, so I just supply a dummy value first and this seems to fix it (the dummy value itself isn't displayed).
p + scale_x_date(labels = c("x", "12/13", "13/14", "14/15", "15/16", "16/17", "17/18", "18/19"))

# to show specific dates only
p + scale_x_date(breaks = as.Date(c("2022-07-01", "2022-07-03", "2022-07-05")))

# sometimes it's difficult to see why axis breaks don't match up with labels. In this case it can be 
# useful to save the graph as an object and then inspect its breaks:
layer_scales(graph_object)$x$break_positions()
```


## Limits and ranges of axes
It's often desirable to alter the range of axes on a graph, e.g. to start an axis at zero. Different options result in different effects - specifically, `limits` will actually cut off data points, whereas other options like `coord_cartesian` will leave the underlying data alone but just display the part of the axis you've specified. 
```{r}
#| eval: false
# coord_cartesian(xlim = c(1,1), ylim = c(1,1)) # (doesn't clip data)
# scale_x_continuous(limits = c(0, 100))
```

`coord_cartesian` alters the visible data.
```{r}
#| eval: true
#| fig-height: 3
#| fig-width: 5
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + geom_point() +
  geom_smooth(aes(colour = Species)) +
  coord_cartesian(xlim = c(4.5, 5.5))
```

`coord_fixed` specifies a ratio for the display of the two axes. A 1-to-1 ratio would be a value of 1. A value <1 compresses the y-axis.
```{r}
#| eval: true
#| fig-height: 3
#| fig-width: 5
# Also try e.g. coord_fixed(20/1), i.e. a 20-to-1 ratio where the x-axis is 20 times
# as long as the y-axis
  
iris %>%
  ggplot(aes(x = Sepal.Length, y = Sepal.Width)) + geom_point() +
  geom_smooth(aes(colour = Species)) +
  coord_fixed(0.5)
```


A tip: if you want a y-axis to begin at zero but don't know the maximum, you can do this:
```{r}
#| eval: false
p + expand_limits(y = 0)
```

When you have a discrete x-axis, sometimes you get a lot of space between aesthetics and the edges of the plot. To control the spacing, add `scale_*_discrete(expand = c(0.1, 0.1))`:
```{r}
#| fig-height: 3
#| fig-width: 5

# regular plot with too much padding:
padding = mpg %>%
    mutate(year = as.factor(year)) %>%
    group_by(year) %>%
    summarise(cty_m = mean(cty, na.rm = T)) %>%
    pivot_longer(c(cty_m)) %>%
    ggplot(aes(x = year, y = value)) +
    stat_summary(fun = "mean", geom = "line", aes(group = 1))
padding

# getting rid of padding:
padding +  scale_x_discrete(expand = c(0.1, 0.1))
```


Clipping determines whether to display elements that would lie outside the plot panel. Expanding sets a buffer margin around a plot to prevent overlapping.
```{r}
#| fig-height: 3
#| fig-width: 5

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 2) +
  # zero expansion
  coord_cartesian(expand = 0,
                  clip = "off") +
  ggtitle("Clipping off")

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(size = 2) +
  # zero expansion
  coord_cartesian(expand = 0,
                  clip = "on") +
  ggtitle("Clipping on")
```

Use custom labels for x-axis: say you have numeric year labels in the data such as `202122` but you want to display these as "2021/22". 
```{r}
#| eval: false

# define a list of values
yr_labels = c("2018/19", "2019/20", "2020/21", "2021/22", "2022/23")

plot + scale_x_continuous(labels = yr_labels)
```

Another option is to use an anonymous function to extract the last two digits of the year
```{r}
plot + scale_x_continuous(breaks = 2013:2024, labels = function(x) substr(as.character(x), 3, 4))
```

If you have a character (discrete) x-axis and want to dodge the labels, try `guide_axis`. Note that this only works if the labels are very short.
```{r}
#| eval: true
df = data.frame(
  x = c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j"),
  y = c(10, 11, 12, 13, 14, 15, 16, 17, 18, 19)
)

df %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() +
  scale_x_discrete(guide = guide_axis(n.dodge = 2))
```


# Themes
To add a pre-defined theme to a plot: `p + theme_*()`.

Tips:

- Place the `theme` argument early on in the plot sequence if you want to adjust other features e.g. axis attributes, because otherwise `theme` will override these. 
- Set a global theme: `theme_set(theme_classic(base_size = 16))`. 

```{r}
#| echo: false
p = mtcars %>% ggplot(aes(x=mpg, y=hp)) + geom_point()

grey = p + theme_grey() + theme(axis.title = element_blank()) + ggtitle("theme_grey")
bw = p + theme_bw() + theme(axis.title = element_blank()) + ggtitle("theme_bw")
line = p + theme_linedraw() + theme(axis.title = element_blank()) + ggtitle("theme_linedraw")
minimal = p + theme_minimal() + theme(axis.title = element_blank()) + ggtitle("theme_minimal")
classic = p + theme_classic() + theme(axis.title = element_blank()) + ggtitle("theme_classic")
void = p + theme_void() + theme(axis.title = element_blank()) + ggtitle("theme_void")

ggarrange(grey, bw, line, minimal, classic, void)
```


## Analysis function theme (`afcharts`)
The AF has created its own package for producing accessible plots, based on ggplot2. The main function is `theme_af()`. It has the following arguments:
```{r}
#| eval: false
theme_af(
  base_size = 14,
  base_line_size = base_size/24,
  base_rect_size = base_size/24,
  grid = c("y", "x", "xy", "none"),
  axis = c("x", "y", "xy", "none"),
  ticks = c("xy", "x", "y", "none"),
  legend = c("right", "left", "top", "bottom", "none")
)
```

See this guidance: "There's general guidelines for research report figures here https://www.gov.uk/government/publications/research-reports-guide-and-template but there's also a package {afcharts} that formats ggplots in line with government guidelines (https://best-practice-and-impact.github.io/afcharts/). It's also recommended to save plots as .svg files as they're editable outside of R".

```{r}
afcharts::use_afcharts()
  
palmerpenguins::penguins %>%
  ggplot(aes(x = island, fill = sex)) +
  geom_bar(position = position_dodge(),
           width = 0.8) +
theme_af(legend = "bottom")
  
  
gapminder %>%
  filter(country %in% c("United Kingdom", "China", "Togo", "Bangladesh")) %>%
  ggplot(aes(x = year, y = lifeExp, colour = country)) +
  geom_line(linewidth = 1) +
  theme_af(legend = "bottom") +
  scale_colour_discrete_af() +
  scale_y_continuous(limits = c(0, 82),
                     breaks = seq(0, 80, 20),
                     expand = c(0, 0)) +
  scale_x_continuous(breaks = seq(1952, 2007, 5)) +
   labs(
    x = "Year",
    y = NULL)
```

You will need to reset the default ggplot2 theme if you don't want to use `afcharts` any more.
```{r}
ggplot2::theme_set(theme_grey())
```


## A custom theme I like
```{r}
#| fig-height: 3
#| fig-width: 5

# make this into a custom function that can be applied to any plot
theme_chris = function(){
  theme(
    axis.line = element_line(colour = "grey80"),
    panel.grid.major.y = element_line(colour = "grey90"),
    panel.grid.major.x = element_blank(), 
    panel.background = element_rect(fill = "white", colour = NA))
}

p + theme_chris()
```


# Legends
`ggplot` will create a legend automatically if any aesthetic is mapped to a linetype, fill, or colour. 
```{r}
#| fig-height: 3
#| fig-width: 5
# by default, legends appear on the right
p = mpg %>%
  mutate(year = as.factor(year),
         cyl = as.factor(cyl)) %>%
  group_by(year, cyl) %>%
  summarise(mean_cty = mean(cty)) %>% 
  ggplot(aes(x = year, y= mean_cty, fill = cyl)) +
  scale_x_discrete(breaks = c(1999, 2008)) +
  geom_bar(stat = "identity", position = position_dodge()) 

p
```


```{r}
#| eval: false
#| fig-height: 3
#| fig-width: 5
# To change:

p + theme(legend.position = "none") 
p + theme(legend.position = "top") 
p + theme(legend.position = "bottom")

# custom location for legend (first number is left-right, second is bottom-top)
p + theme(legend.position = c(0.5, 0.9))
```


To control legend wrapping (i.e. a single horizontal row instead of wrapping into two rows) - especially when `legend.position = "bottom"`.
```{r}
#| eval: true
#| fig-height: 3
#| fig-width: 5
p + theme(legend.position = "bottom") + guides(fill = guide_legend(nrow = 1)) 
```

Filling and outlining:
```{r}
#| eval: true
#| fig-height: 3
#| fig-width: 5

# make box transparent
p + theme(legend.background = element_blank(),
          legend.position = c(0.9, 0.7))

# add grey box
p + theme(legend.box.background = element_rect(colour = "grey", linewidth = 2))

# fill box
p + theme(legend.background = element_rect(fill = "grey"))
```


You can control the palette, breaks, labels, and name. E.g. if the factor labels are too long, you can shorten them. Use a separate call to contol other attributes like `linetype` or `fill`. In fact, the `scale_*_manual` functions are incredibly versatile for this kind of task (and are discussed in more detail below).
```{r}
#| fig-height: 3
#| fig-width: 5
gcookbook::countries %>%
    filter(str_detect(Name, "United")) %>%
    ggplot(aes(x = Year, y = GDP)) + 
    geom_line(aes(colour = Name), linewidth = 0.8) +
    scale_colour_manual(values = c(RColorBrewer::brewer.pal(3, "Set2")),
                        breaks = c("United Kingdom", "United Arab Emirates", "United States"),
                        labels = c("UK", "UAE", "US"),
                        name = "Country") +
  theme_minimal()
```

## Direct labels
Sometimes a legend isn't necessary or the right aesthetic choice. Instead, it's possible to append labels directly to dots or lines using `directlabels` or `ggrepel`.
```{r}
#| echo: false
# data to use
df = structure(list(time_period = c(200910, 201011, 201112, 201213, 
201314, 201415, 201516, 201617, 201718, 201819, 201920, 202021, 
202122, 202223, 200910, 201011, 201112, 201213, 201314, 201415, 
201516, 201617, 201718, 201819, 201920, 202021, 202122, 202223, 
200910, 201011, 201112, 201213, 201314, 201415, 201516, 201617, 
201718, 201819, 201920, 202021, 202122, 202223), subject = c("Drama", 
"Drama", "Drama", "Drama", "Drama", "Drama", "Drama", "Drama", 
"Drama", "Drama", "Drama", "Drama", "Drama", "Drama", "Media/Film/TV", 
"Media/Film/TV", "Media/Film/TV", "Media/Film/TV", "Media/Film/TV", 
"Media/Film/TV", "Media/Film/TV", "Media/Film/TV", "Media/Film/TV", 
"Media/Film/TV", "Media/Film/TV", "Media/Film/TV", "Media/Film/TV", 
"Media/Film/TV", "Music", "Music", "Music", "Music", "Music", 
"Music", "Music", "Music", "Music", "Music", "Music", "Music", 
"Music", "Music"), total_exam_entries = c(80765, 74755, 70371, 
69754, 70984, 70851, 67890, 62112, 57176, 57659, 58070, 57126, 
53585, 49568, 58586, 51783, 49478, 48698, 51328, 51602, 46830, 
40843, 37912, 30494, 29052, 27468, 25483, 26988, 45433, 43157, 
40761, 41256, 42446, 43698, 41650, 38897, 34709, 34580, 34672, 
35451, 33893, 29870)), row.names = c(NA, -42L), class = c("tbl_df", 
"tbl", "data.frame"))
```

Using `directlabels` and `geom_dl`. Note that other method options include `first.points` and `last.qp` which adjusts the size of the text automatically). This also requires clipping to be turned off and for the plot margins to be extended (otherwise text won't be displayed properly).
```{r}
#| fig-height: 3.5
#| fig-width: 5
df %>%
  ggplot(aes(x = time_period, y = total_exam_entries)) +
  geom_line(aes(colour = subject), linewidth = 0.8) +
  ylab("Entries") + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle=270, vjust = 0.5),
        legend.position = "none",
        axis.line = element_line(colour="black"),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::label_comma(big.mark = ","),
                     limits = c(0,80000)) +
  scale_x_continuous(breaks = c(200910, 201011, 201112, 201213, 201314, 201415, 201516, 201617, 
                                201718, 201819, 201920, 202021, 202122, 202223),
                     labels = c("2009/10", "2010/11", "2011/12", "2012/13", "2013/14", "2014/15",
                                "2015/16", "2016/17", "2017/18", "2018/19", "2019/20", "2020/21",  
                                "2021/22", "2022/23")) +
  geom_dl(aes(label=subject), method=list("last.points", "bumpup", cex=0.8)) +
  coord_cartesian(clip="off") +
  theme(plot.margin = unit(c(1,4,1,1), "lines")) 
```

Using `geom_label_repel`. Note that this requires a bit more wrangling to make sure only the final points are displayed (using the `data` argument) - otherwise every single point will be labelled.
```{r}
#| fig-height: 3.5
#| fig-width: 5

df %>%
  ggplot(aes(x = time_period, y = total_exam_entries, label = subject)) +
  geom_line(aes(colour = subject), linewidth = 0.8) +
  ylab("Entries") + 
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle=270, vjust = 0.5),
        legend.position = "none",
        axis.line = element_line(colour="black"),
        panel.grid.minor = element_blank()) +
  scale_y_continuous(labels = scales::label_comma(big.mark = ","),
                     limits = c(0,80000)) +
  scale_x_continuous(breaks = c(200910, 201011, 201112, 201213, 201314, 201415, 201516, 201617, 
                                201718, 201819, 201920, 202021, 202122, 202223),
                     labels = c("2009/10", "2010/11", "2011/12", "2012/13", "2013/14", "2014/15",
                                "2015/16", "2016/17", "2017/18", "2018/19", "2019/20", "2020/21",  
                                "2021/22", "2022/23")) +
  coord_cartesian(clip = "off") +
  geom_label_repel(aes(label = subject), 
                   label.padding = .15, 
                   data = df %>% group_by(subject) %>% filter(time_period == max(time_period)),
                   size = 3, hjust = 0.5, nudge_x=0.5)

```


# Other things
## Flip coordinates
```{r}
#| fig-height: 3
#| fig-width: 5

ggplot2::diamonds %>%
  ggplot(aes(x=cut, y=carat)) +
  geom_violin() +
  coord_flip()
```



# Using colours
## Manual
Quick manual palettes for sequential and categorical variables (adjust to number needed).
```{r}
seq2 = c("#12436D", "#6BACE6")
scales::show_col(seq2)

seq2 = c("#4B1D73", "#C1A0D7")
scales::show_col(seq2)

seq2 = c("#0C4C5A", "#5DC1B9")
scales::show_col(seq2)

seq2 <- c("#7A0C2E", "#E87B79")
scales::show_col(seq2)

seq2 <- c("#37474F", "#90A4AE")
scales::show_col(seq2)

seq3 = c("#12436D", "#2073BC", "#6BACE6") 
scales::show_col(seq3)

cat4 = c("#12436D", "#28A197", "#801650", "#F46A25", "#3D3D3D", "#A285D1")
scales::show_col(cat4)
```


```{r}
# Sequential palettes
blues = c("#104F75", "#407291", "#7095AC", "#9FB9C8", "#CFDCE3")
reds = c("#8A2529", "#A15154", "#B97C7F", "#D0A8A9", "#E8D3D4")
oranges = c("#E87D1E", "#ED974B", "#F1B178", "#F6CBA5", "#FAE5D2")
yellows = c("#C2A204", "#CEB536", "#DAC768", "#E7DA87", "#F3ECCD")
greens = c("#004712", "#336C41", "#669171", "#99B5A0", "#CFDABD")
purples = c("#260859", "#51397A", "#7D6B9B", "#A89CBD", "#D4CEDE")

show_col(c(blues, reds, oranges, yellows, greens, purples), ncol=5, cex_label=0.7)
```


## `scale_*_manual`
```{r}
#| fig-height: 3
#| fig-width: 6
# example of sequential. Tip: use rev() to reverse the order of colours
ggplot2::diamonds %>%
  filter(price < 6000) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) +
  scale_fill_manual(values = rev(yellows))
```

An example of how to specify manual linetype and colour. Note that these must use the same breaks in order to be presented as a single legend (as opposed to separate legends for linetype and for colour):
```{r}
#| eval: false
scale_linetype_manual(name = "",
                        values = c("solid", "dashed"),
                        breaks = c("pay", "median_after_tax"),
                        labels = c("Pay", "Median UK")) +
scale_colour_manual(name = "",
                        values = c("black", "red"),
                        breaks = c("pay", "median_after_tax"),
                        labels = c("Pay", "Median UK"))
```

If you would like NA to be a different colour by default
```{r}
#| eval: false
plot +
  scale_fill_discrete(na.value = "white")
```


Use a focus palette to bring attention to one particular series, while keeping the other data there for comparison.
```{r}
#| fig-height: 3
#| fig-width: 5

focus = c("#12436D", "#BFBFBF", "#BFBFBF") # contrast ratio of 5.57:1 (passes Web Content Accessibility
# Guidelines [WCAG]).
# Add as many grey colours as you need.

countries %>% 
  filter(str_detect(Name, "United")) %>%
  ggplot(aes(x=Year, y=GDP)) + 
  geom_line(aes(colour = Name), size = 0.8) +
  scale_colour_manual(values = focus,
                        breaks = c("United Kingdom", "United Arab Emirates", "United States"),
                        labels = c("UK", "UAE", "US"),
                        name = "Country")
```

Alternatively, use `gghighlight`:
```{r}
#| fig-height: 3
#| fig-width: 5

countries %>% 
  filter(str_detect(Name, "Republic")) %>%
  ggplot(aes(x=Year, y=GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  gghighlight(str_detect(Name, "Czech"),
              label_params = list(size = 3),
              label_key = Code,
              unhighlighted_colour = "grey70")
```


## Other colour scale options
```{r}
#| eval: true
#| fig-height: 3
#| fig-width: 5

# if a fill or colour variable is continuous, the default scale will be scale_*_continuous,
# so there's no need to specify it directly:
p = gcookbook::heightweight %>%
  ggplot(aes(x = heightIn, y = weightLb, colour = heightIn)) +
  geom_point(size = 2)

# note: scale_colour_continuous works within R but causes a fatal error when
# knitting to html, so replace with scale_colour_gradient

p + scale_colour_gradient(na.value = "grey50") +
  theme_void() +
  ggtitle("scale_colour_gradient (default)")

p + scale_colour_gradient(low = "white", high = "red", na.value = "black") +
  theme_void() +
  ggtitle("scale_colour_gradient")

# scale_colour_gradient2 allows you to set a mid-point colour but you have to specify this from
# the dataframe. Of course you don't have to use the median as the 'midpoint' here. 
p + scale_colour_gradient2(low = "blue", mid = "white", high = "red",
  midpoint = median(gcookbook::heightweight$heightIn)) +
  theme_void() +
  ggtitle("scale_colour_gradient2")

# colour_gradientn allows you to specify your own set of colours for a whole spectrum.
# note: if you don't specify a vector of values, the colours will be evenly positioned along
# the scale. However, you can control this manually for instance if you want to highlight the very 
# top values. Just specify a vector of values between 0 and 1 to correspond to how you want the 
# colours mapped to the scale, e.g. values = c(1, 0.9, 0.8, 0.7, 0)
p + scale_colour_gradientn(colours = c("red","yellow","green","lightblue","darkblue"),
                                    values = c(1, 0.9, 0.8, 0.7, 0)) +
  theme_void() +
  ggtitle("scale_colour_gradientn")

```

NOTE! If you want to map to integers like 18, 21, 22, 23, 39 then you need to rescale these from 0 to 1, which the `scales` package can do for you:
```{r}
#| eval: false

p + scale_colour_gradientn(colours = c("red","yellow","green","lightblue","darkblue"),
                                    values = scales::rescale(c(39, 23, 22, 21, 18), to = c(0,1)),
                                    name = "Legend name") +
  theme_void() 
```


## Brewer palettes
`scale_*_brewer` to use predefined palettes. See a list of available palettes [here](https://jmsallan.netlify.app/blog/the-brewer-palettes/). 

### Brewer monochrome palettes:

* Blues
* Greens
* Purples
* Greys
* Oranges
* Reds

```{r}
#| fig-height: 3
#| fig-width: 6

ggplot2::diamonds %>%
  filter(price <6000) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) + 
  scale_fill_brewer(palette = "Greens") +
  theme_void()
```
### Brewer spectral palettes
* BuGn
* BuPu
* GnBu
* OrRed
* PuBu
* PuRd
* RdPu
* YlGn
* PuBuGn
* YlGnBu
* YlOrBr
* YlOrRd

```{r}
#| fig-height: 3
#| fig-width: 6

ggplot2::diamonds %>%
  filter(price <6000) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) + 
  scale_fill_brewer(palette = "YlOrRd") +
  theme_void()
```

### Brewer diverging palettes

* Spectral
* RdYlGn
* RdYlBu
* RdGy
* RdBu
* PuOr
* PRGn
* PiYG
* BrBG

```{r}
#| fig-height: 3
#| fig-width: 6

ggplot2::diamonds %>%
  filter(price <6000) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) + 
  scale_fill_brewer(palette = "Spectral") +
  theme_void()
```
### Brewer qualitative palettes
* Accent
* Dark2
* Paired
* Pastel1 and Pastel2
* Set1
* Set2
* Set3

```{r}
#| fig-height: 3
#| fig-width: 6

ggplot2::diamonds %>%
  filter(price <6000) %>%
  ggplot(aes(x = price, fill = cut)) +
  geom_histogram(position = "dodge", binwidth = 1000) + 
  scale_fill_brewer(palette = "Set2") +
  theme_void()
```


Example of a categorical palette from the [Analysis Function](https://analysisfunction.civilservice.gov.uk/support/communicating-analysis/introduction-to-data-visualisation-e-learning/data-visualisation-e-learning-module-3-accessibility-and-colour-palettes/). 
```{r}
# note: recommendation is to use a max of 4 colours
categorical = c("#12436D", "#28A197", "#801650", "#F46A25", "#3D3D3D", "#A285D1")
show_col(categorical)
```

How to use a custom palette:
```{r}
#| fig-height: 3
#| fig-width: 5  
ggplot2::midwest %>%
  filter(county %in% c("ADAMS", "ALEXANDER", "BOND", "BOONE", "BROWN")) %>%
  ggplot(aes(x = county, y = poptotal, fill = county)) +
  geom_col(position = position_dodge()) +
  scale_fill_manual(values = categorical)
```


Using `scale_fill_brewer` palettes:
```{r}
#| fig-height: 3
#| fig-width: 5

ggplot2::midwest %>%
  filter(county %in% c("ADAMS", "ALEXANDER", "BOND", "BOONE", "BROWN")) %>%
  ggplot(aes(x=county, y=poptotal, fill = county)) +
  geom_col(position = position_dodge()) +
  scale_fill_brewer(type = "seq", palette = "Set1")
```


# Mapping
Use the `sf` package and a shapefile to visualise data on a map. This shapefile comes from the UK government's [Geoportal](https://geoportal.statistics.gov.uk/search?q=BDY_RGN%3BDEC_2023&sort=Title%7Ctitle%7Casc). This [tool](https://borders.ukdataservice.ac.uk/bds.html) is also a good resource for finding different shapefiles. 
```{r}
library(sf)
library(tidyverse)

# firstly read in a shapefile. Note that when you download a shapefile, it comes with other 
# files like .shx. These all need to be in the same directory as the .shp file itself.
# As such, you can actually just specify the folder name that contains the shapefile and its
# accompanying files. 

eng_reg_map = st_read("./Shapefiles/Eng_regional_2023/RGN_DEC_2023_EN_BFC.shp")

# check that the map is appropriate for your needs by graphing it without any data attached
eng_reg_map %>%
  ggplot() +
  geom_sf(fill   = "white",
          colour = "black") +
  theme_void()

# next, add some data from a preexisting source:
df = data.frame(
  region = c("North West", "North East", "Yorkshire and The Humber", 
             "East of England", "West Midlands", "East Midlands", 
             "South East", "London", "South West"),
  n = c(81, 25, 39, 49, 51, 43, 53, 115, 32)
)

# join this to the sf object and map. 
# use scale_fill_gradient to control fill colours

eng_reg_map %>%
  left_join(df, by = c("RGN23NM" = "region")) %>%
  ggplot(aes(fill = n)) +
  geom_sf(colour = "grey50") +
  theme_void() +
  scale_fill_gradient(name = "Respondents",
                      low = "grey90", high = "grey15", 
                      na.value = "white")
```

To add points based on geometry:
```{r}
# say you have a data frame from GIAS with easting and northing values of schools (these could be lat/long too)
map_df = data.frame(
                    school = c(1:15),
                    easting = c(427144, 390498, 391459, 453596, 332195, 439952, 397143, 331548, 504305, 527647, 349039, 437349, 504740, 448965, 166988),
                    northing = c(433927, 323814, 162039,  86591, 391270, 114064, 157408, 390872, 208312, 169328, 405788, 377158, 103181, 361343,  44078),
                    pupils  = c(28, 23, 59, 22, 28, 76, 13, 14, 57, 43, 25, 12, 59, 52, 31)
)

# the shapefile is based on long/lat, so you need to convert Eastings and Northings.
# The first step here is to check which CRS the original map uses
st_crs(eng_reg_map)$epsg

# use the st_as_sf function to create another version of your dataframe (map_df) - feed in the crs
# and it will add an additional column called 'geometry'

schools.sf = st_as_sf(map_df, coords = c("easting", "northing"), remove = FALSE, crs = 27700)

# check that the points themselves work
# plot(schools.sf$geometry)

# add to the map and add text labels
eng_reg_map %>%
  ggplot() +
  geom_sf(fill   = NA,
          colour = "grey50") +
  geom_sf(data = schools.sf, colour = "red", size = 2) +
  ggrepel::geom_text_repel(data = schools.sf,
                  aes(x = st_coordinates(schools.sf)[,1],
                      y = st_coordinates(schools.sf)[,2],
                      label = school)) +
  theme_void()
# optionally if you have used aes(colour = pupils): + scale_colour_gradient(low = "grey", high = "red")
```

# Odd bits for sorting later

## Vertical and horizontal lines
```{r}
#| fig-height: 3
#| fig-width: 5
mpg %>% 
  ggplot(aes(x=displ, y = cty)) + 
  geom_point() + 
  geom_vline(xintercept = 4.5, colour = "red", linetype="dashed") +
  geom_hline(yintercept = 22.5, colour = "purple", linetype = "twodash")
```


### Shading
```{r}
#| fig-height: 3
#| fig-width: 5
# geom_rect
ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point() +
     annotate("rect", xmin = 3, xmax = 4.2, ymin = 12, ymax = 21,
        alpha = .2)
  
# to use with dates:
# annotate("rect", xmin = as.Date("2020-06-01"), xmax = as.Date("2021-06-01"), ymin = -Inf, ymax = Inf,
#           alpha = .2)
```


# Dynamic plots
There may be cases where you want to produce multiple plots with the same geoms and structure but looking at a different outcome variable each time. Here it is more efficient to write a single function and repeat it. This can be made dynamic through use of `{{` and `deparse`:
```{r}
#| eval: false
la_graph = function(data, outcome){
  
  # extract title passed to function for using in ggtitle later
  title = deparse(substitute(outcome))
  
  data %>%
    ggplot(aes(x = year, y = {{outcome}})) +
    theme_classic() +
    geom_line(aes(group = la_name), linewidth = 0.7, colour = "grey90") +
    stat_summary(fun = mean, geom = "line", aes(group = la_group, colour = la_group),
                 linewidth = 1) +
    theme(axis.text.x = element_text(angle=45, hjust = 1),
          axis.title.x = element_blank(),
          legend.title = element_blank()) +
    scale_y_continuous(labels = scales::label_percent(accuracy = 1, scale = 100),
                       limits = c(0,1)) +
    scale_x_continuous(breaks = 2013:2024, labels = function(x) substr(as.character(x), 3, 4))  +
    # invoke title object here
    ylab(title) +
    ggtitle(title)
}

# to run:
la_graph(dataset, outcome_of_interest)
```



# Combining and saving plots
## Facetting
To facet by a single variable, use `facet_wrap`. Add `scales = "free"` if axes are different for each group (or try `free_x` or `free_y`). NOTE: the old way was `~ species`, but `vars` allows you to specify several groupings, e.g. `vars(species, sex)`. 
```{r}
#| fig-height: 5
#| fig-width: 6
palmerpenguins::penguins %>% 
  ggplot(aes(x = bill_length_mm, y = flipper_length_mm)) + 
  geom_point(aes(colour = island)) +
  facet_wrap(vars(species), 
             nrow = 3,
             scales = "fixed") +
  # customisation of strips/panels:
  theme(strip.text = element_text(face = "bold", color = "black", hjust = 0.5, size = 10),
        strip.background = element_rect(fill = NA, linetype = "solid", color = "black")) +
  theme_bw()
```

To facet by two variables, use `facet_grid`
```{r}
#| fig-height: 5
#| fig-width: 6
palmerpenguins::penguins %>% 
  drop_na(sex) %>%
  ggplot(aes(x=bill_length_mm, y=flipper_length_mm)) + 
  geom_point() +
  facet_grid(species ~ sex,
             scales = "fixed") +
  theme_bw()
```


## Arranging
Use `ggarrange` from the `ggpubr` package to arrange plots.
```{r}
#| fig-height: 3.5
#| fig-width: 7
# plot 1
a = gcookbook::countries %>%
    filter(Name == "United Kingdom") %>%
    ggplot(aes(x = Year, y = GDP)) + 
    geom_line() +
    scale_y_continuous(labels = scales::label_comma(big.mark = ",")) +
    theme_bw()

# plot 2
b = gcookbook::countries %>%
    filter(Name == "United Kingdom") %>%
    ggplot(aes(x = Year, y = healthexp)) + 
    geom_line() +
    scale_y_continuous(labels = scales::label_comma(big.mark = ",")) +
    theme_bw()

# combine
ggarrange(a, b, 
          ncol=2, 
          widths = c(1,1),
          labels = c("A", "B"))
```

You can also nest ggarrange calls, e.g. if you have 3 or more plots
```{r}
#| fig-width: 7
c = gcookbook::countries %>%
    filter(Name == "United Kingdom") %>%
    ggplot(aes(x = Year, y = infmortality)) + 
    geom_line() +
    theme_bw()

ggarrange(a,
          ggarrange(b, c, ncol = 2),
          nrow = 2)
```

It's also possible to use a shared legend and title (if applicable to all plots)
```{r}
#| fig-width: 8
a = gcookbook::countries %>%
    filter(str_detect(Name, "United")) %>%
    ggplot(aes(x = Year, y = GDP)) + 
    geom_line(aes(colour = Name, linetype = Name), linewidth = 0.8) +
    theme_bw()

# plot 2
b = gcookbook::countries %>%
    filter(str_detect(Name, "United")) %>%
    ggplot(aes(x = Year, y = healthexp)) + 
    geom_line(aes(colour = Name, linetype = Name), linewidth = 0.8) +
    theme_bw()

# wrap in annotate_figure for common title
annotate_figure(ggarrange(a, b, ncol=2,
          common.legend = T,
          legend = "bottom"),
    top = text_grob("A common title", color = "red", face = "bold", size = 14),
    bottom = text_grob("Data source: \n Countries data set", color = "blue",
                                  hjust = 1, x = 0.99, face = "italic", size = 10),
    fig.lab.pos = "top",
    fig.lab.size = 14)
```

Another recipe for a common legend. This comes from the English Hubs impact eval where there were 6 individual plots combined into a matrix plot. Each had its own legend which was inefficient. This recipe required one plot with all waves together so as to get a legend with all available values. Then a custom function is used to extract the legend from this. Aftwards, combine the individual plots with the extracted legend. Note that the recipe below is only useful if you want to make many pairwise comparisons, e.g. compare some level of a factor against some constant (in this case, compare control schools against each wave separately). 
```{r}
#| eval: false
# create plot with all values on it together
all = wave_plot_all(y1_schl)

# extract the legend from this
# from here: https://www.geeksforgeeks.org/r-language/add-common-legend-to-combined-ggplot2-plots-in-r/
get_only_legend <- function(plot) {
  plot_table <- ggplot_gtable(ggplot_build(plot))
  legend_plot <- which(sapply(plot_table$grobs, function(x) x$name) == "guide-box")
  legend <- plot_table$grobs[[legend_plot]]
  return(legend)
}

legend = get_only_legend(all)

# create the individual plots (this also uses a custom plotting function)
w1 = wave_plot_single(y1_schl, wave = 201920, "#12436D", "Wave 1", treatment_start = "2019-09-01")
w2 = ...

# combine these
combined_plot = grid.arrange(w1, w2, w3, w4, w5, w6)

# combine this with the extracted legend
grid.arrange(combined_plot, legend, nrow = 2, heights = c(10, 1))
```

Building on this further, if you would like to add a figure caption then there this is this rather convoluted example using the grid package and arrangeGrob. The below is a reproducible example.
```{r}
#| eval: true

# make plot with all
all = gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "United States", "Spain", "Paraguay",
                     "India")) %>%
  ggplot(aes(x = Year, y = GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  theme_bw() +
  scale_colour_manual(name = NULL,
                      values = c("red", "#12436D", "#28A197", "#801650", "#F46A25"),
                      breaks = c("United Kingdom", "United States", "Spain", "Paraguay",
                                 "India")) +
  theme(legend.position = "bottom",
        legend.text = element_text(size=11),
        legend.background = element_rect(colour = "grey", linewidth = 1)) +
  guides(colour = guide_legend(nrow = 1)) 
all

# extract legend
get_only_legend <- function(plot) {
  plot_table <- ggplot_gtable(ggplot_build(plot))
  legend_plot <- which(sapply(plot_table$grobs, function(x) x$name) == "guide-box")
  legend <- plot_table$grobs[[legend_plot]]
  return(legend)
}

legend = get_only_legend(all)

# note: looks like ggpubr::get_legend can do this already
legend = ggpubr::get_legend(all)

# create individual plots
p1 = gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "United States")) %>%
  ggplot(aes(x = Year, y = GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(name = "Country",
                      values = c("red", "#12436D"),
                      breaks = c("United Kingdom", "United States")) 

p2 = gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "Spain")) %>%
  ggplot(aes(x = Year, y = GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(name = "Country",
                      values = c("red", "#28A197"),
                      breaks = c("United Kingdom", "Spain")) 

p3 = gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "Paraguay")) %>%
  ggplot(aes(x = Year, y = GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(name = "Country",
                      values = c("red", "#801650"),
                      breaks = c("United Kingdom", "Paraguay"))

p4 = gcookbook::countries %>%
  filter(Name %in% c("United Kingdom", "India")) %>%
  ggplot(aes(x = Year, y = GDP)) + 
  geom_line(aes(colour = Name), linewidth = 0.8) +
  theme_bw() +
  theme(legend.position = "none") +
  scale_colour_manual(name = "Country",
                      values = c("red", "#F46A25"),
                      breaks = c("United Kingdom", "India")) 

# combine
combined_plot = gridExtra::grid.arrange(p1, p2, p3, p4)

# combine the individual plots, legend, a title, and a caption:
main_plot <- gridExtra::arrangeGrob(combined_plot, legend, 
                         nrow = 2, 
                         heights = c(10, 1.5))

title <- grid::textGrob("Main Title for Combined Plots",
                  gp = grid::gpar(fontsize = 16, fontface = "bold"),
                  hjust = 0.5, x = 0.5)

caption <- grid::textGrob("Note: This is some explanatory text.",
                    gp = grid::gpar(fontsize = 11),  # font size
                    hjust = -0.05,                 # left-align
                    x = 0)

final_plot <- gridExtra::arrangeGrob(title, main_plot, caption,
                          ncol = 1,
                          heights = c(0.1, 1, 0.05))

grid::grid.newpage()
grid::grid.draw(final_plot)
```


## Saving
Files will be saved in your current working directory.
```{r}
#| eval: false
my_plot = ggplot(...)

ppi = 300  # pixels per inch
png("file_name.png", width = 4*ppi, height = 4*ppi, res = ppi)
my_plot
dev.off()
```


## Looping or iterating
Sometimes you want to produce the same graph for a range of different levels in a factor. This recipe below produces histograms, grouped by 'EAL' for every variable listed in the `covariates` vector. It uses `purrr::map` to iterate through this list and then invokes `cowplot::plot_grid` to add these all to one file.
```{r}
#| eval: false
hist_group_fun = function(x, y = NA) {
  ggplot(data, aes(x = .data[[x]], fill = .data[[y]], colour = .data[[y]]) ) +
    geom_histogram(alpha=0.5, position = "identity") +
    theme(legend.text = element_text(size=9))
}

hist_group_fun(x = "IDACI.rank19", y = "EAL")

# make a list of variables to loop through
covariates = c("cpm.raw", "bpvs.raw", "IDACI.rank19")

# iterate through covariates
covariate_plots = map(covariates, ~ hist_group_fun(.x, "EAL") )

# save to a single pdf file
pdf("histograms_all.pdf")
cowplot::plot_grid(plotlist = covariate_plots)
dev.off()
```
  

Another way to achieve this is by converting variable names to symbols. The function below creates dodged bar plots (with the same y-axis value each time), and then the map function iterates through a list of grouping variables (`vars`) such that each time a new plot is produced with the next grouping factor in the vector. The function uses tidy evaluation (the unquote or 'bang-bang' character `!!`)
```{r}
#| eval: false
bar_fun = function(data, var1, var2) {
  
  # convert grouping factors into symbols first
  var1 = sym(var1)
  var2 = sym(var2)
  
  data %>%
    filter(psc_pass == "Fail") %>%
    # the below effectively becomes group_by(c_summer_born, SEX)
    group_by(!!var1, !!var2) %>%
    summarise(PHONICSMARK = round(mean(PHONICSMARK, na.rm = TRUE), 1), .groups = "drop") %>%
    ggplot(aes(x = !!var1, y = PHONICSMARK, fill = !!var2)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    geom_text(aes(label = PHONICSMARK), 
              vjust = 1.2, 
              position = position_dodge(width = 1),
              show.legend = T) +
    labs(title = paste("By", as_string(var2))) +
    theme_classic() +
    theme(axis.title = element_blank())
}

vars = c("SEX", "c_disad", "c_any_Sen", "c_white_british", "c_eal")
  
plots = purrr::map(vars, ~ bar_fun(data = ph_cen2, var1 = "c_summer_born", var2 = .x))

patchworK::wrap_plots(plots, ncol = 3)

```

This slightly different recipe takes an outcome variable and grouping factor to produce a histogram. It uses `enquo` to extract the number of levels in the specified grouping factor which it then supplies to the `nrow` argument of `facet_wrap`.
```{r}
#| eval: false
dist_graph = function(data, x, char){
  
  # quote the 'char' variable
  char_var = enquo(char)
  
  # find unique number of values in 'char' by unquoting
  width = data %>% 
    summarise(dist = n_distinct(!!char_var)) %>% as.numeric()
  
  data %>%
    ggplot(aes(x = {{x}}, fill = {{char}})) +
    geom_histogram(binwidth = 1, position  = "identity") +
    facet_wrap(vars({{char}}), 
               scales = "free",
               nrow = width) +
    theme(legend.position = "none",
          axis.title.y = element_blank()) +
    geom_vline(xintercept = 32, colour = "red")
}

dist_graph(c1_c2, x = PHONICSMARK_y1, char = MoB)
```

