---
title: "Regression Modelling Workflow"
date: now
date-format: "[Last updated:] DD/MM/YYYY HH:mm"
format:
  html:
    standalone: true
    embed-resources: true
    toc: true
    toc-depth: 2
    toc-expand: 4
    number-sections: true
    toc-location: left-body
    html-table-processing: none
    df-print: kable
    code-fold: false
    theme: default
    fontsize: 1em
    link-external-newwindow: true
execute:
  warning: false
  echo: true
  eval: false
---

# Introduction
This document sets out a template workflow for the regression modelling process, including linear and logistic applications.  

# Step 1: Prepare Data

- Determine the dependent variable: interpreted in raw units or scaled? (Remember that a DV doesn't have to show linearity for linear regression)
- Determine relevant covariates 
    - E.g. FSM, SEN, Sex, Ethnicity, Summer-born, prior attainment, school type, region
    - Useful to make into factors and prefix with `c_`:

```{r}
  mutate(c_fsm6 = if_else(fsm6 == 1, 1, 0),
         c_sen = if_else(senf == "N", 0, 1),
         c_eal = if_else(flang == 1, 1, 0),
         c_male = if_else(sex == "M", 1, 0),
         c_white_british = if_else(ethnic == "WBRI", 1, 0),
         c_summer_born = if_else(monthofbirth %in% c(5:8), 1, 0))
```





# Step 2: Interrogate Data

- Descriptive statistics for all variables
- Correlations for numeric variables

- How does the distribution of the DV differ according to each covariate?

```{r}
# a matrix of plots of every variable against one another
# numeric variables will show scatterplots and histograms;
# discrete variables will show boxplots with other numeric and bar charts/facetbars
# with other discrete
df %>%
  select(...) %>%
  GGally::ggpairs(upper = list(continuous = "smooth", combo = "box_no_facet", discrete = "count", na = "na"),
                  lower = list(continuous = "smooth", combo = "facethist", discrete = "facetbar", na = "na"))
```


```{r}
# for a linear DV:
# produces a multi-panel density plot for each covariate
df %>%
  select(DV, starts_with("c_")) %>%
  pivot_longer(-c(DV)) %>%
  mutate(value = as.factor(value)) %>%
  ggplot(aes(x = DV, fill = factor(value))) +
  geom_density(bw=1, alpha=0.5) +
  facet_wrap(vars(name), scales="free")
```


# Step 3: Run First Model

Quick-start options:

- `lm()` - linear DV, single time point
- `glm()` - binary DV, single time point
- `lme4::lmer()` - linear DV, multilevel (e.g. repeated measures, clustered)
- `lme4::glmer()` - binary DV, multilevel (e.g. repeated measures, clustered)

```{r}
# linear model

m = lm(DV ~ x1 + x2*x3 +
         c_1 +
         c_2 +
         c_3, 
       data = df)

# multilevel model
m = lme4::lmer(DV ~ x1 + x2*x3 + (1 | pmr),
            na.action = na.omit, 
            REML = F,  
            control = lmerControl(optimizer = "bobyqa", calc.derivs = T), 
            data = df)
```

```{r}
# logistic model

m = glm(DV ~ x1 + x2*x3,
        data = df,
        family = binomial(link = "logit"))

m = lme4::glmer(DV ~ x1 + x2*x3 + (1 | pmr),
            na.action = na.omit, 
            REML = F, 
            family = binomial,
            control = glmerControl(optimizer = "bobyqa"),
            data = df)
```

Next, check the model summary output:

```{r}
summary(m) # linear
broom::tidy(m) # linear
broom::tidy(m, exponentiate = T) # logistic
broom::glance(m) # model fit stats
```



# Step 4: Run Diagnostics



# Step 5: Amend Model


# Step 6: Reassess, compare fit, rerun diagnostics



# Step 7: 